---
title: "Comprehensive Guide to MCMCDiag"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comprehensive Guide to MCMCDiag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
library(MCMCDiag)
```

Introduction
The MCMCDiag package provides tools for diagnosing the convergence of Markov Chain Monte Carlo (MCMC) simulations. This vignette demonstrates how to use the package to perform convergence diagnostics, including the Gelman-Rubin diagnostic, trace plots, and effective sample size (ESS) calculations.

MCMC methods are widely used in Bayesian statistics to sample from posterior distributions. Assessing the convergence of these methods is crucial for reliable inference. The MCMCDiag package implements several diagnostic tools to help users evaluate the convergence of their MCMC chains.

Installation

You can install the MCMCDiag package from GitHub using the following commands:
```{r}
# Install devtools if not already installed
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# Install MCMCDiag from GitHub
devtools::install_github("abinalikunju/MCMCDiag")
```


Once installed, load the package:

```{r}
library(MCMCDiag)
```

{r}
Copy code

Basic Usage
Let's start by generating some sample MCMC chains for demonstration purposes:

{r}
Copy code
set.seed(123)
n_chains <- 3
n_iterations <- 1000
chains <- lapply(1:n_chains, function(x) rnorm(n_iterations))
Gelman-Rubin Diagnostic
The Gelman-Rubin diagnostic assesses convergence by comparing within-chain and between-chain variances:

{r}
Copy code
gr_stat <- gelman_rubin(chains)
print(gr_stat)
A value close to 1 indicates good convergence.

Trace Plots
Visualize the chains using trace plots:

{r}
Copy code
trace_plot(chains[[1]], "Chain 1")
Enhanced Trace Plot
For multiple chains, it’s important to ensure each layer includes the required aesthetic mappings:

{r}
Copy code
enhanced_trace_plot(chains, c("Chain 1", "Chain 2", "Chain 3"))
Effective Sample Size
Calculate the effective sample size:

{r}
Copy code
ess <- calculate_ess(chains)
print(ess)
Advanced Topics
Handling Large Datasets
MCMCDiag is designed to handle large datasets efficiently. Here's an example with longer chains:

{r}
Copy code
large_chains <- lapply(1:3, function(x) rnorm(1e5))
system.time(gr_stat_large <- gelman_rubin(large_chains))
Efficient Processing of Long Chains
For very long chains, you might want to thin the chains before analysis:

{r}
Copy code
thin_chains <- lapply(large_chains, function(chain) chain[seq(1, length(chain), by = 10)])
system.time(gr_stat_thin <- gelman_rubin(thin_chains))
Function Reference
gelman_rubin()

Calculates the Gelman-Rubin diagnostic statistic.

{r}
Copy code
?gelman_rubin
trace_plot()

Generates a trace plot for a single chain.

{r}
Copy code
?trace_plot
enhanced_trace_plot()

Generates trace plots for multiple chains.

{r}
Copy code
?enhanced_trace_plot
calculate_ess()

Calculates the effective sample size.

{r}
Copy code
?calculate_ess
mcmc_diag()

Performs comprehensive MCMC diagnostics.

{r}
Copy code
diag_result <- mcmc_diag(chains)
str(diag_result)
print(diag_result)
summary(diag_result)
plot(diag_result)  # Ensure that plot method is correctly implemented to handle the output
Conclusion
The MCMCDiag package provides essential tools for diagnosing the convergence of MCMC chains. By using the Gelman-Rubin diagnostic, trace plots, and effective sample size calculations, users can ensure that their MCMC simulations have converged and can be used for reliable Bayesian inference.
