---
title: "Using MCMCDiag for MCMC Convergence Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCMCDiag}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```


## Introduction

The MCMCDiag package provides tools for diagnosing the convergence of posterior distributions generated using MCMC methods. This vignette demonstrates how to use the package to perform convergence diagnostics, including the Gelman-Rubin diagnostic, trace plots, and effective sample size (ESS) calculations. These tools help assess whether MCMC chains have converged and can be used for reliable inference.

## Loading the Package
Make sure the MCMCDiag package is installed and load it using the following code:

### Install from GitHub

```{r}
devtools::install_github("abinalikunju/MCMCDiag")
```


Load the package

```{r setup}
library(MCMCDiag)
```


### Generating Sample Data
Code for generating sample MCMC chains:

```{r}

set.seed(123)
n_chains <- 3
n_iterations <- 1000
chains <- lapply(1:n_chains, function(x) rnorm(n_iterations))

```


### Gelman-Rubin Diagnostic

The Gelman-Rubin diagnostic assesses convergence by comparing within-chain and between-chain variances:


```{r} 

gr_stat <- gelman_rubin(chains)
print(gr_stat)

```

### Enhanced Trace Plot

Visualize multiple chains with the enhanced trace plot:

```{r}

p <- enhanced_trace_plot(chains, c("Param1", "Param2", "Param3"))
print(p)

```


### Effective Sample Size

Calculate the effective sample size:

```{r }

ess <- calculate_ess(chains)
print(ess)

```

### Geweke Diagnostic

The Geweke diagnostic compares means from different portions of the chain:

```{r }

geweke_stat <- geweke_diagnostic(chains[[1]])
print(geweke_stat)

```

### Heidelberger-Welch Test

Perform the Heidelberger-Welch stationarity test:

```{r}

hw_test <- heidelberger_welch_test(chains[[1]])
print(hw_test)

```

### MCMC Diagnostics

Use the mcmc_diag function for a comprehensive diagnostic:

```{r}

diag_result <- mcmc_diag(chains)
print(diag_result)
plot(diag_result)
summary(diag_result)

```

### Conclusion

The MCMCDiag package provides essential tools for diagnosing the convergence of MCMC chains. By calculating the Gelman-Rubin statistic and generating trace plots, the package also aims to compute the effective sample size, users can ensure that their MCMC chains have converged, enabling accurate Bayesian inference. Additionally, the package aims to handle large datasets and long chains efficiently.
